|

### 2024-09-05 manuscript/paperCMSACversion.Rmd
added references

While any player on a hockey team can get into a fight, enforcers are exposed to many more fights than the typical NHL player. As a result of being exposed to additional contact, they are potentially exposed to additional head trauma beyond the amount associated with playing hockey. There are many questions about the long-term implications of this head trauma and its potential to result in negative health outcomes in the future. The ingrained nature of fighting into the game of hockey is greatly criticized and there are many active debates about its current declining trend. Figure \ref{fig:enforcer_density} shows a decline in the proportion of enforcers to non-enforcers as players get younger. The question is, should the game continue to move away from contact-driven strategy in favor of younger, faster, highly-skilled players?

```{r enforcer_density, echo = FALSE, fig.cap='Density Plot of Study Cohorts by DOB'}
cohort_dat = dat %>% mutate(cohort = case_when(PIM.GP > 3 ~ 'Enforcer',
                                  Total.Fights > 50 ~ 'Enforcer',
                                  .default = 'Non-Enforcer'))

@@

ggplot(cohort_dat, aes(x = DOB)) + geom_density(aes(fill = cohort),alpha = 0.75 )
```

@@

Based on their analysis (a matched cohort study) they find that there is no statistically significant difference in the mortality rates between enforcers and controls for either definition of enforcer, but they do conclude that enforcers died on average about 10 years earlier than their non-enforcer counterparts. In addition, they draw the conclusion that there is not a significant difference in all-cause mortality rate but find enforcers died more frequently of drug overdose and suicide.

@@

```{r KMEF, echo = FALSE, fig.cap="Kaplan Meier plot for EF" }
library(ggsurvfit)
survfit2(Surv ~ EFchar, data = dat) %>%
  ggsurvfit() +
  labs(
    x = "Age",
    y = "Overall survival probability"
  ) +
  add_confidence_interval()
```

```{r KMEP, echo = FALSE, fig.cap="Kaplan Meier plot for EP" }
library(ggsurvfit)
survfit2(Surv ~ EPchar, data = dat) %>% 
  ggsurvfit() +
  labs(
    x = "Age",
    y = "Overall survival probability"
  ) + 
  add_confidence_interval()
```

@@

Figures \ref{fig:KMEF} and \ref{fig:KMEP} show Kaplan-Meier plots for EF and EP.  There is substantial overlap in the K-M curves between the enforcer and non-enforcers groups when enforcers is defined using the EF criteria. However, the enforcer group when defined by the EP criteria exhibits evidence of lower survival probabilities starting around age 50 relative to the non-enforcer group, though it is important to remember that these two groups differ substantially in age, height, and weight.

In order to account for these differences betwen the groups, Cox proportional hazards models @CoxPH were fit: one for each definition of enforcer defined in @Popkin2023 (i.e. E-F and E-P) with time until death as the outcome variable.  In addition, the indicators for each definition of enforcer, games played (GP), height (Ht), weight (Wt), year of birth (DOB), and position indicators (Center (C), Wing (W), or Defender (D)) were added to each model as control variables.

@@

```{r popkin_conclusions, echo = FALSE, tab.cap = ""}
cf = read.csv('../JAMA NHL Data/C-F-Table 1.csv')
cp = read.csv('../JAMA NHL Data/C-P-Table 1.csv')

@@

```
<!-- latex table generated in R 4.4.1 by xtable 1.8-4 package
Wed Sep  4 18:44:57 2024-->

\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline
 & enforcer & mean\_YOB & mean\_age\_at\_death \\ 
  \hline
1 & 0.00 & 1958.29 & 56.46 \\ 
  2 & 1.00 & 1962.81 & 47.15 \\ 
   \hline
\end{tabular}
\caption{Original Paper Conclusions Data Summary
 \label{tab:popkin_conclusions}}
\end{table}

These results suggest that the conclusions of @Popkin2023 are overstated. We believe this may be due to the differences in exposure time between the groups at the time of their concluding analysis. Table \ref{tab:popkin_conclusions} shows that by conditioning on death, the mean year of birth becomes significantly different between enforcer vs non-enforcer groups. Before conditioning on death, the mean year of birth for each group was `r round(precnclsn_sum[[1,2]], 2)` for non-enforcers and `r round(precnclsn_sum[[2,2]], 2)` for enforcers. Conditioning on death contradicts any previous equality of means due to matching, resulting in the enforcer group having an average year of birth much later than the control group, and thereby it is no longer appropriate to compare their mortality rate BECAUSEEEE.....

@@

Talk about the conditioning on death stuff.

Age difference statement in Popkin is over stated.

>>>>

While any player on a hockey team can get into a fight, enforcers are exposed to many more fights than the typical NHL player. As a result of being exposed to additional contact, they are potentially exposed to additional head trauma beyond the amount associated with playing hockey. There are many questions about the long-term implications of this head trauma and its potential to result in negative health outcomes in the future. The ingrained nature of fighting into the game of hockey is greatly criticized and there are many active debates about its current declining trend. Figure \ref{fig:enforcer_density} shows a decline in the proportion of enforcers to non-enforcers as players  were born more recently. The question is, should the game continue to move away from contact-driven strategy in favor of younger, faster, highly-skilled players?

```{r enforcer_density, echo = FALSE, fig.cap='Density Plot of Study Cohorts by DOB'}
cohort_dat = dat %>% mutate(cohort = case_when(PIM.GP > 3 ~ 'Enforcer',
                                  #Total.Fights > 50 ~ 'Enforcer',
                                  .default = 'Non-Enforcer'))

@@

cohort_dat$cohort <- factor(cohort_dat$cohort, levels = c('Non-Enforcer', 'Enforcer'))

#ggplot(cohort_dat, aes(x = DOB)) + geom_density(aes(fill = cohort),alpha = 0.75 )

ggplot(cohort_dat, aes(x = DOB)) +
  geom_freqpoly(aes(color = cohort),alpha = 0.75 , position = 'stack') +
  scale_color_manual(values = c('#00bFC4', '#F8766d'))
```

@@

Based on this matched cohort study, they conclude that there is no statistically significant difference in the mortality rates between enforcers and controls for either definition of enforcer. However, they do find that enforcers died on average about 10 years earlier than their non-enforcer counterparts and more frequently of drug overdose and suicide.

@@

```{r KMEF, echo = FALSE, fig.cap="Kaplan Meier plot for E-F" }
library(ggsurvfit)
survfit2(Surv ~ EFchar, data = dat) %>%
  ggsurvfit() +
  labs(
    x = "Age",
    y = "Overall survival probability"
  ) +
  add_confidence_interval()
```

```{r KMEP, echo = FALSE, fig.cap="Kaplan Meier plot for E-P" }
library(ggsurvfit)
survfit2(Surv ~ EPchar, data = dat) %>% 
  ggsurvfit() +
  labs(
    x = "Age",
    y = "Overall survival probability"
  ) + 
  add_confidence_interval()
```

@@

Figures \ref{fig:KMEF} and \ref{fig:KMEP} show Kaplan-Meier plots for E-F and E-P.  There is substantial overlap in the K-M curves between the enforcer and non-enforcers groups when enforcers is defined using the EF criteria. However, the enforcer group when defined by the EP criteria exhibits evidence of lower survival probabilities starting around age 50 relative to the non-enforcer group, though it is important to remember that these two groups differ substantially in age, height, and weight.

In order to account for these differences between the groups, Cox proportional hazards models (@CoxPH) were fit: one for each definition of enforcer defined in @Popkin2023 (i.e. E-F and E-P) with time until death as the outcome variable.  In addition, the indicators for each definition of enforcer, games played (GP), height (Ht), weight (Wt), year of birth (DOB), and position indicators (Center (C), Wing (W), or Defender (D)) were added to each model as control variables.

@@

<!--
```{r popkin_conclusions, echo = FALSE, tab.cap = ""}
cf = read.csv('../JAMA NHL Data/C-F-Table 1.csv')
cp = read.csv('../JAMA NHL Data/C-P-Table 1.csv')

@@

```
 latex table generated in R 4.4.1 by xtable 1.8-4 package
Wed Sep  4 18:44:57 2024

These results suggest that the conclusions of @Popkin2023 are overstated. We believe this may be due to the differences in exposure time between the groups at the time of their concluding analysis. Table \ref{tab:popkin_conclusions} shows that by conditioning on death, the mean year of birth becomes significantly different between enforcer vs non-enforcer groups. Before conditioning on death, the mean year of birth for each group was `r round(precnclsn_sum[[1,2]], 2)` for non-enforcers and `r round(precnclsn_sum[[2,2]], 2)` for enforcers. Conditioning on death contradicts any previous equality of means due to matching, resulting in the enforcer group having an average year of birth much later than the control group, and thereby it is no longer appropriate to compare their mortality rate BECAUSEEEE.....
-->

@@

<!--Talk about the conditioning on death stuff.

Age difference statement in Popkin is over stated.  -->
